---
title: "SDS230_FinalProject"
author: "Tejita Agarwal, Lelan Hu, Franklin She"
date: "5/5/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages
library(plyr)

```

**Introduction**
Background and motivation, not more than a short paragraph

**Data**
*The data used in this analysis are:

1. school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira); categorical
2. sex - student's sex (binary: 'F' - female or 'M' - male); categorical
3. age - student's age (numeric: from 15 to 22); continuousm (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
4. studytime - weekly study time (numeric: 1 - 10 hours); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
5. higher - wants to take higher education (binary: yes or no); categorical
6. Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
7. Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
8. health - current health status (numeric: from 1 - very bad to 5 - very good); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
9. absences - number of school absences (numeric: from 0 to 93); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
10. grade - final grade (numeric: from 0 to 20)*; continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)

**Data Cleaning Process**
Describe the cleaning process you used on your data. Talk about what issues you encountered.
Demonstrate use of find/replace, data cleaning, dealing with missing values, text character replacement, matching. It’s ok if your data didn’t require much of this.

```{r}
all_data <- read.csv("student-por.csv", header=TRUE)
# decided to just use one of the csv files and not merging cause it wasn't working, but if you guys can figure it out pls. but it think just using one is fine.
selected_data <- all_data[, c("school", "sex", "age", "studytime", "higher", "Dalc", "Walc", "health", "absences", "G3")]

# rename G3 to grade
selected_data <- rename(selected_data, c("G3"="grade"))

# remove observations where grade is 0
selected_data <- selected_data[selected_data$grade != 0,]

nrow(selected_data)
names(selected_data)
head(selected_data)

attach(selected_data)
# i think the data already comes pretty clean tbh... 
```


**Descriptive Plots**
Plots should be clearly labeled, well formatted, and display an aesthetic sense.
Show appropriate use of at least ONE of each of the following – boxplot, scatterplot (can be matrix plot), normal quantile plot (can be related to regression), residual plots, histogram.
```{r}
boxplot(grade ~ school, main = "Boxplot of Grades by School", ylab = "Grade / Scale(0 - 20)", col = "blue", xlab = "School")
boxplot(grade ~ sex, main = "Boxplot of Grades by Sex", ylab = "Grade / Scale(0 - 20)", col = "green", xlab = "Sex")

##plot(age, Walc)
plot(jitter(age, factor = .5), jitter(Walc), pch = 19, col = "red", xlab = "Age",
     ylab = "Weekend Alcohol Consumption")
mtext("Weekend Alcohol Consumption vs. Age", cex = 1.2, line = 1)
mtext(paste("Sample Correlation =", round(cor(age, Walc), 3)), cex = 1.2, line = 0)

hist(grade, main = "Histogram of Grades", xlab = "Grade (0 - 20)", xlim = c(0, 20), col = "orange")
# grade is pretty much the only variable that is normally distributed.. is this an issue?

library(PerformanceAnalytics)
chart.Correlation(selected_data[, c(3,4,6:10)], histogram = TRUE, pch = 19)
##might be useful to transform Walc, Dalc, and absences. for now, I will try a log transformation

transdata <- selected_data[, c(3,4,6:10)]
transdata$logWalc <- log(transformeddata$Walc)
transdata$logDalc <- log(transformeddata$Dalc)
transdata$logabs <- log(transformeddata$absences)
chart.Correlation(transdata[, c(1,2,5,7:10)], histogram = TRUE, pch = 19)
##does anyone know why it says NaN in the last column for the correlations with logabs

```

**Summary Information**

**Analysis**
Basic tests - t-test, correlation, AND ability to create bootstrap confidence interval for either a t-test or a correlation.
Permutation Test – include at least one.
Multiple Regression – use either backwards stepwise regression or some form of best subsets regression. Should include residual plots. A GLM with a mix of continuous and categorical predictors is fine here.
AT LEAST ONE OF THE FOLLOWING TECHNIQUES – ANOVA, ANCOVA, Logistic Regression, Multinomial Regression, OR data scraping off a website.
```{r}
t.test(Walc ~ sex) ##
t.test(log10(Walc) ~ sex) ## still significant
t.test(studytime ~ sex)
t.test(grade ~ sex)
t.test(grade ~ school)
t.test(Walc ~ school) ## not significant
t.test(log10(Walc) ~ school) ## still not significant

##correlations are already provided in the matrix plot above, should we do this again?

##someone please do bootstrap of something here





```


**Conclusions**
a short paragraph
##remember to take our names off!!
