---
title: "SDS230_FinalProject"
author: "Tejita Agarwal, Lelan Hu, Franklin She"
date: "5/5/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages
library(plyr)

```

**Introduction**
Background and motivation, not more than a short paragraph

**Data**
*The data used in this analysis are:

1. school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
2. sex - student's sex (binary: 'F' - female or 'M' - male)
3. age - student's age (numeric: from 15 to 22)
4. studytime - weekly study time (numeric: 1 - 10 hours)
5. higher - wants to take higher education (binary: yes or no)
6. Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
7. Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
8. health - current health status (numeric: from 1 - very bad to 5 - very good)
9. absences - number of school absences (numeric: from 0 to 93)
10. grade - final grade (numeric: from 0 to 20)*

**Data Cleaning Process**
Describe the cleaning process you used on your data. Talk about what issues you encountered.
Demonstrate use of find/replace, data cleaning, dealing with missing values, text character replacement, matching. It’s ok if your data didn’t require much of this.

```{r}
all_data <- read.csv("student-por.csv", header=TRUE)
# decided to just use one of the csv files and not merging cause it wasn't working, but if you guys can figure it out pls. but it think just using one is fine.
selected_data <- all_data[, c("school", "sex", "age", "studytime", "higher", "Dalc", "Walc", "health", "absences", "G3")]

# rename G3 to grade
selected_data <- rename(selected_data, c("G3"="grade"))

# remove observations where grade is 0 or 1
selected_data <- selected_data[selected_data$grade > 1,]

nrow(selected_data)
names(selected_data)
head(selected_data)

# i think the data already comes pretty clean tbh... 
```


**Descriptive Plots**
Plots should be clearly labeled, well formatted, and display an aesthetic sense.
Show appropriate use of at least ONE of each of the following – boxplot, scatterplot (can be matrix plot), normal quantile plot (can be related to regression), residual plots, histogram.
```{r}
hist(selected_data$grade)
# grade is pretty much the only variable that is normally distributed.. is this an issue?
```



**Summary Information**

**Analysis**
Basic tests - t-test, correlation, AND ability to create bootstrap confidence interval for either a t-test or a correlation.
Permutation Test – include at least one.
Multiple Regression – use either backwards stepwise regression or some form of best subsets regression. Should include residual plots. A GLM with a mix of continuous and categorical predictors is fine here.
AT LEAST ONE OF THE FOLLOWING TECHNIQUES – ANOVA, ANCOVA, Logistic Regression, Multinomial Regression, OR data scraping off a website.

**Conclusions**
a short paragraph
