---
title: "SDS230_FinalProject"
author: "Tejita Agarwal, Lelan Hu, Franklin She"
date: "5/5/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Packages
library(plyr)

```

**Introduction**
Background and motivation, not more than a short paragraph

**Data**
*The data used in this analysis are:

1. school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira); categorical
2. sex - student's sex (binary: 'F' - female or 'M' - male); categorical
3. age - student's age (numeric: from 15 to 22); continuousm (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
4. studytime - weekly study time (numeric: 1 - 10 hours); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
5. higher - wants to take higher education (binary: yes or no); categorical
6. freetime - free time after school (numeric: from 1 - very low to 5 - very high); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
7. goout - going out with friends (numeric: from 1 - very low to 5 - very high); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
8. Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
9. Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
10. health - current health status (numeric: from 1 - very bad to 5 - very good); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
11. absences - number of school absences (numeric: from 0 to 93); continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)
12. grade - final grade (numeric: from 0 to 20)*; continuous (discrete integers, makes sense to be treated as continuous variable, because of the nature of the variable)

**Data Cleaning Process**
Describe the cleaning process you used on your data. Talk about what issues you encountered.
Demonstrate use of find/replace, data cleaning, dealing with missing values, text character replacement, matching. It’s ok if your data didn’t require much of this.

```{r}
all_data <- read.csv("student-por.csv", header=TRUE)
# decided to just use one of the csv files and not merging cause it wasn't working, but if you guys can figure it out pls. but it think just using one is fine.
selected_data <- all_data[, c("school", "sex", "age", "studytime", "higher", "freetime", "goout", "Dalc", "Walc", "health", "absences", "G3")]

# rename G3 to grade
selected_data <- rename(selected_data, c("G3"="grade"))

# remove observations where grade is 0
selected_data <- selected_data[selected_data$grade != 0,]

nrow(selected_data)
names(selected_data)
head(selected_data)

attach(selected_data)
# i think the data already comes pretty clean tbh... 
```


**Descriptive Plots**
Plots should be clearly labeled, well formatted, and display an aesthetic sense.
Show appropriate use of at least ONE of each of the following – boxplot, scatterplot (can be matrix plot), normal quantile plot (can be related to regression), residual plots, histogram.
```{r}
# boxplots of grade by school and by sex
boxplot(grade ~ school, main = "Boxplot of Grades by School", ylab = "Grade / Scale(0 - 20)", col = "blue", xlab = "School")
boxplot(grade ~ sex, main = "Boxplot of Grades by Sex", ylab = "Grade / Scale(0 - 20)", col = "green", xlab = "Sex")

# scatterplots of age and alcohol consumption (weekend and weekday)
plot(jitter(age, factor = 1), jitter(Walc), pch = 19, col = "red", xlab = "Age",
     ylab = "Weekend Alcohol Consumption", cex = 0.5)
mtext("Weekend Alcohol Consumption vs. Age", cex = 1.2, line = 1)
mtext(paste("Sample Correlation =", round(cor(age, Walc), 3)), cex = 1.2, line = 0)


plot(jitter(age, factor = 1), jitter(Dalc), pch = 19, col = "blue", xlab = "Age",
     ylab = "Weekday Alcohol Consumption", cex = 0.5)
mtext("Weekday Alcohol Consumption vs. Age", cex = 1.2, line = 1)
mtext(paste("Sample Correlation =", round(cor(age, Dalc), 3)), cex = 1.2, line = 0)


# histogram of grades
hist(grade, main = "Histogram of Grades", xlab = "Grade (0 - 20)", xlim = c(0, 20), col = "orange")
hist(freetime, main = "Histogram of Freetime", xlab = "Freetime)", col = "orange", breaks = 6)
hist(goout, main = "Histogram of Going Out", xlab = "Time Going Out With Friends)", col = "orange")
# TODO: tej can u fix the correlation charts to include the new variables (freetime, goout)
# they are the 6th and 7th variable of selected data, other data is moved up


library(PerformanceAnalytics)
chart.Correlation(selected_data[, c(3,4,6:12)], histogram = TRUE, pch = 19)
##might be useful to transform Walc, Dalc, and absences. for now, I will try a sqrt transformation of abs, and log transformation of Walc and Dalc. 


transdata <- selected_data[, c(3,4,6:12)]
transdata$logWalc <- log(transformeddata$Walc + 1)
transdata$logDalc <- log(transformeddata$Dalc + 1)
transdata$sqrtabs <- sqrt(transformeddata$absences)
chart.Correlation(transdata[, c(1:4,7,9:12)], histogram = TRUE, pch = 19)

## dalc should be none or some
## for future analysis, it might be preferable to recode Dalc to indicate binary: do they drink on weekdays yes/no.

```

**Summary Information**

**Analysis**
Basic tests - t-test, correlation, AND ability to create bootstrap confidence interval for either a t-test or a correlation.
Permutation Test – include at least one.
Multiple Regression – use either backwards stepwise regression or some form of best subsets regression. Should include residual plots. A GLM with a mix of continuous and categorical predictors is fine here.
AT LEAST ONE OF THE FOLLOWING TECHNIQUES – ANOVA, ANCOVA, Logistic Regression, Multinomial Regression, OR data scraping off a website.
```{r}
t.test(Walc ~ sex) ##
t.test(log10(Walc) ~ sex) ## still significant
t.test(studytime ~ sex)
t.test(grade ~ sex)
t.test(grade ~ school)
t.test(Walc ~ school) ## not significant
t.test(log10(Walc) ~ school) ## still not significant

##correlations are already provided in the matrix plot above, should we do this again?

##someone please do bootstrap of something here


# Multiple Regression to predict grade (Best Subsets Regression, Bayesian Information Criteria)
mod1 <- regsubsets(grade ~ ., data = selected_data, nvmax = 11)
mod1sum <- summary(mod1)
names(selected_data)[mod1sum$which[which.min(mod1sum$bic), ]][-1] # names of the predictors
gradeBIC <- selected_data[,mod1sum$which[which.min(mod1sum$bic), ]]
modfin <- lm(grade ~ ., data = gradeBIC)
summary(modfin) # summary of best model according to BIC
# TODO: pls check my work ^^^ ... it seems one of the chosen predictors (freetime) has a p-value greater than 0.05?


```


**Conclusions**
a short paragraph
##remember to take our names off!!

