---
title: "SDS230_FinalProject"
author: "Tejita Agarwal, Lelan Hu, Franklin She"
date: "5/5/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Introduction**
Background and motivation, not more than a short paragraph

**Data**
Make a LIST of all variables you actually use – describe units, anything I should know. Ignore variables you don’t discuss.

**Data Cleaning Process**
Describe the cleaning process you used on your data. Talk about what issues you encountered.
Demonstrate use of find/replace, data cleaning, dealing with missing values, text character replacement, matching. It’s ok if your data didn’t require much of this.

```{r}
d1=read.csv("student-mat.csv",header=TRUE)
d2=read.csv("student-por.csv",header=TRUE)
d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet"))

print(nrow(d3))

```


**Descriptive Plots**
Plots should be clearly labeled, well formatted, and display an aesthetic sense.
Show appropriate use of at least ONE of each of the following – boxplot, scatterplot (can be matrix plot), normal quantile plot (can be related to regression), residual plots, histogram.


**Summary Information**

**Analysis**
Basic tests - t-test, correlation, AND ability to create bootstrap confidence interval for either a t-test or a correlation.
Permutation Test – include at least one.
Multiple Regression – use either backwards stepwise regression or some form of best subsets regression. Should include residual plots. A GLM with a mix of continuous and categorical predictors is fine here.
AT LEAST ONE OF THE FOLLOWING TECHNIQUES – ANOVA, ANCOVA, Logistic Regression, Multinomial Regression, OR data scraping off a website.

**Conclusions**
a short paragraph
